---
title: "Chapter 6: Analysis of longitudinal data"
output: html_document
---

#  Chapter 6: Analysis of longitudinal data

## Loading the datasets

```{r}

BPRSL <- read.csv("/Users/suvi/Documents/GitHub/IODS-project/Data/bprsl.csv", header = TRUE, sep = ",")

RATSL <- read.csv("/Users/suvi/Documents/GitHub/IODS-project/Data/ratsl.csv", header = TRUE, sep = ",")
```

## 1. Analysis of RATSL dataset

### Graphical display

```{r}

library(ggplot2)

ggplot(RATSL, aes(x = WD, y = rats, linetype = as.factor(ID))) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ as.factor(Group), labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$rats), max(RATSL$rats)))

```

There are differences between groups as can be seen from the upper picture. Let's see what happens if we standardize the data and draw the plot again.

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

RATSLS <- RATSL %>%
  group_by(WD) %>%
  mutate(stdrats = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

glimpse(RATSLS)

ggplot(RATSLS, aes(x = WD, y = stdrats, linetype = as.factor(ID))) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ as.factor(Group), labeller = label_both) +
  scale_y_continuous(name = "standardized rats")

```

Differences between groups are still clear when pictures are drawn with standardized rats.

### Summary measure

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

n <- RATSL$WD %>% unique() %>% length()

RATSLU <- RATSL %>%
  group_by(Group, WD) %>%
  summarise( mean = mean(rats), se = sd(rats/sqrt(n) )) %>%
  ungroup()

glimpse(RATSLU)

ggplot(RATSLU, aes(x = WD, y = mean, linetype = as.factor(Group), shape = as.factor(Group))) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(rats) +/- se(rats)")

```


The plot shows the mean response profiles for every group. As can be seen, group one is different from the other groups. 

### Find out the outlier


```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

RATSLfilt <- RATSL %>%
  filter( WD > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

glimpse(RATSLfilt)

ggplot(RATSLfilt, aes(x = as.factor(Group), y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), groups")
```

Here is a boxplot of the mean versus groups. 

I create a new dataset by filtering the outlier and then I'll try to draw the same boxplot again.

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

RATSLfilt2 <- RATSLfilt %>%
  filter(mean > 300)

ggplot(RATSLfilt2, aes(x = as.factor(Group), y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), groups")

```

I filtered the data with mean>300 and group 1 disappeared. So there is only groups 2 and 3 and the boxplot shows the rats means versus those groups.

### Anova test

```{r}

RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = T)

library(ggplot2)
library(dplyr)
library(tidyr)

t.test(mean ~ Group, data = RATSLfilt2, var.equal = TRUE)

RATSAN <- RATSLfilt %>%
  mutate(baseline = RATS$WD1)

fit <- lm(mean ~ baseline + as.factor(Group), data = RATSAN)

anova(fit)

```

As can be seen from the Anova test result, the group difference is not statistically significant. 


## 2. Analysis of BPRSL dataset

Data is already in and it is in a long format model. I'll change the variables treatment and subject to factors.

```{r}

library(dplyr)

BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

glimpse(BPRSL)
```

### Graphical display

```{r}

library(dplyr)
library(tidyr)

dim(BPRSL)

ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```


### The linear model

Next I will fit a multiple linear regression model with bprs as response and week and treatment as explanatory variables.

```{r}

library(dplyr)
library(tidyr)

BPRSL_reg <- lm(bprs ~ week + treatment, data = BPRSL)

summary(BPRSL_reg)
```

If we look at the p-values, week has a small value which indicates statistical significance. Treatment2 has a p-value bigger than 0,05. 

### The random intercept model

Next I will fit the random intercept model for the same two explanatory variables. 

```{r}

library(dplyr)
library(tidyr)
library(lme4)

BPRSL_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

summary(BPRSL_ref)
```

Fitting a random intercept model allows the linear regression fit for each individual. T-value looks to be -10 with week variable, which indicates statistical significance.

### Random intercept and random slope model

Next I'm able to fit the random intercept and random slope model to the dataset. Fitting a random intercept and random slope model allows the linear regression fits for each individual to differ in intercept but also in slope. 


```{r}

library(dplyr)
library(tidyr)
library(lme4)
library(ggplot2)

BPRSL_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRSL_ref1)

anova(BPRSL_ref1, BPRSL_ref)
```



### Random intercept and random slope model with interaction

Finally, I can fit a random intercept and slope model that allows for a treatment × week interaction.

```{r}

library(dplyr)
library(tidyr)
library(lme4)
library(ggplot2)

BPRSL_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRSL_ref2)

anova(BPRSL_ref2, BPRSL_ref1)

ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Weeks", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Bprs") +
  theme(legend.position = "top")

Fitted <- fitted(BPRSL_ref2)

BPRSL <- BPRSL %>%
  mutate(Fitted)

ggplot(BPRSL, aes(x = week, y = Fitted, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Weeks", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Bprs") +
  theme(legend.position = "top")

```




